Model and results for identity signaling project

### 2020-06-16 paper/ms.tex
updated paper pushing towards fresh draft with current results

\vspace{1em}
\begin{table}[H]
  \centering
  \begin{tabular}{cl}
    Global parameter & Description \\
    \toprule 
    $K$      & Number of relevant traits determining similarity, liking, disliking, etc. \\
    $w$      & Homophily \\
    $d$      & Interaction penalty when one agent dislikes the other \\
    $\delta$ & Additional interaction penalty when both agents dislike each other \\
    $s$      & Added interaction benefit when agents are similar, as defined by their traits \\
    $S$      & Number of traits two agents must have in common to be considered similar \\
    $M$      & Number of traits used (if any) to define majority/minority populations \\
  \end{tabular}
  \caption{Global model parameters.}
  \label{tab:params}
\end{table}

@@

Model individuals are \emph{agents} who 
\begin{enumerate}
  \item have static ``in-born'' traits that determine pairwise similarity or dissimilarity
  \item hold attitudes about all other agents in the system
  \item have adopted a signaling and receiving strategy
\end{enumerate}Signaling strategies
may be either \emph{covert} or \emph{overt}, and receiving strategies may be
either \emph{generous} or \emph{churlish}. Unless otherwise noted, agents are
randomly initialized to be covert or overt with equal probability. The 
proportion of agents with specific signaling or receiving strategies will be
set to different values in the invasion experiments, explained in more detail below.

@@

\subsubsection{Design}

All computational experiments we implemented share a core design. The only 
difference between experiments is how we initialize agents and which
outcome measures we analyze. This basic design initializes $N$ agents with
a trait vector, an attitude vector, and a signaling and receiving strategy.
Agent attributes can be set fully randomly, as we do for the basic experiments focusing
on the role of key parameters of homophily ($w$), disliking penalty ($d$), and 
covert signaling efficiency ($r$). Randomly setting agent attributes means
choosing $\pm 1$ for each element of agent trait vectors with equal 
probabilities, and choosing a signaling and receiving strategy at random
with equal probabilities for each strategy type. Attributes are initialized
according to the agent's receiving strategy. Generous agents have all 
attitudes about other agents set to 0, indicating neutrality. Churlish agents
have all attitudes set to -1, since these agents dislike others by default.
This and all experiments iterate the model $T$ times, which
means $T$ signaling rounds, with $I$ interactions between signaling rounds,
and social learning after the signaling rounds. In this basic experiment,
we analyze the global prevalence of covert signalers at time $T$ ($\rho_{cov,T}$)
and the global prevalence of generous receivers at time $T$ ($\rho_{gen,T}$),
including calculating correlation between the two.

The first variation on this basic model is intended to test our 
hypothesis that covert signaling prevalence in 
minority populations ($\rho_{cov}^{minor}$) is inversely correlated with
homophily. For this, we initialize the first $M$ traits out of $K$ to mark the 
minority and majority populations. The minority population has its first
$M$ trait vector elements set to $+1$ and the majority population has its
first $M$ trait vector elements set to $-1$. For example, if $K=5$, $M=2$,
and agent $i$ is in the minority, then its trait vector would be
$\tau_i = (-1, -1, -1, \pm 1, \pm 1)$; if $i$ is in the majority, its 
trait vector would be $\tau_i = (1, 1, 1, \pm 1, \pm 1)$, where in the last
two positions the value is chosen randomly with equal probabilities.
In this experiment we analyze the prevalence of covert signalers and 
generous receivers in the minority and majority populations at time T: 
$\rho_{cov,T}^{minor}$, $\rho_{cov,T}^{major}$,
$\rho_{gen,T}^{minor}$, and $\rho_{gen,T}^{major}$.

(NEXT WRITE INVASION PAR)

\subsubsection{Implementation}

@@

@@

@@

Our computational results support this theoretical claim. As the heatmaps
below reveal... In the SI we show regressions of $\rho_{cov,T} \sim \rho_{gen,T}$.

@@

@@

\subsection{Non-signaling as an adaptive strategy}

>>>>

\vspace{1em}
\begin{table}[H]
  \centering
  \begin{tabular}{cp{3.5in}c}
    Symbol & Definition & Values tested (default in bold)  \\
    \toprule 
    $w$      & \emph{Homophily}, the ability of agents to preferentially interact with similar others & $\{0.0, 0.05, 0.10, \ldots, 0.5\}$ \\
    $d$      & \emph{Disliking penalty} imposed when one agent dislikes the other & $\{0.0, 0.05, 0.1,\ldots \mathbf{0.25}, \ldots, 0.5\}$ \\
    $\delta$ & Additional penalty when pair dislike one another & \textbf{0.25} \\
    $E_{cov}$ & \emph{Efficiency of covert signaling} & $\{0.0, 0.05, 0.1,\ldots \mathbf{0.25}, \ldots, 0.5\}$ \\
    $E_{ov}$ & \emph{Efficiency of overt signaling} & \textbf{0.5} \\
    $s$      & \emph{Similarity benefit} received by similar interacting pair & \textbf{??} \\
    $K$      & Number of agent identity traits & $\{\mathbf{3}, 9\}$ \\
    $S$      & Minimum fraction of traits in common for a pair to be considered similar & 
      $\{0.1, 0.2, \ldots, \mathbf{0.5}, \ldots, 1.0\}$\\
    $M$      & Number of traits used to define majority/minority populations & $\{\mathbf{2}, 4\}$\\
  \end{tabular}
  \caption{Global, as opposed to agent-specific, model parameters. 
  Explanatory model parameters are used to answer
  our guiding questions about identity communication strategies. In the final
  column the default value is in bold, if there is one. Homophily never has a
  default value because it is varied in every computational experiment
  condition we test.}
  \label{tab:params}
\end{table}

@@

Model individuals are \emph{agents} who 
\begin{enumerate}
  \item have static ``in-born'' traits that determine pairwise similarity or dissimilarity
  \item hold attitudes about all other agents in the system
  \item have adopted a signaling and receiving strategy
\end{enumerate}
Signaling strategies
may be either \emph{covert} or \emph{overt}, and receiving strategies may be
either \emph{generous} or \emph{churlish}. Unless otherwise noted, agents are
randomly initialized to be covert or overt with equal probability. The 
proportion of agents with specific signaling or receiving strategies will be
set to different values in the invasion experiments, explained in more detail below.

@@

Using the basic components of Agents with attributes and communication, interaction,
and learning processes, we simulated the time course of prevalence of 
covert/overt signalers ($\rho_{cov/ov,t}$) and churlish/generous receivers 
($\rho_{ch/gen,t}$) among the simulated
population as a whole or among agents specified as minorities or majorities.
To understand when and why different communication strategies become prevalent
over others, we varied a number of parameters with important real-world
interpretations that answer this question. Specifically, we want to know
the effects of varying \emph{explanatory parameters} 
homophily ($w$), covert signaling efficiency ($r$), and the
payoff penalty when one interaction partner dislikes the other ($d$). 
(EXPLAIN HOW THESE ANSWER OUR RESEARCH QUESTIONS). We analyze these questions
in three different framings.

Our experiments are divided into three main parts: (1) a basic experiment
to understand the role of the explanatory parameters in the evolution of
communication strategies; (2) an experiment where
we assign all agents to be in the
minority or majority to understand how minorities utilize covert signaling;
and (3) ``invasion'' (BETTER WORD?) experiments to understand how different
communication strategies spread. In our first experiment, we assume nothing 
about the distribution of agent traits, $\tau_{ik}$, other than each is 
drawn randomly from $\{-1, 1\}$. Similarly, all communication strategies
are randomly initialized with each strategy given equal probability of
being chosen. In our second experiment we create majorities and minorities
by setting $M$ of $K$ $\tau_{ik}$ to be
$-1$ for some majority of agents and $+1$ for some minority of agents, where
the fraction of agents in the minority is denoted $\rho^{minor}$. 
In our third and final experiment, we examine what conditions allow for the
spread of different communication strategies when they are initially rare,
which is called an invasion analysis, borrowing from evolutionary theory.
For one example of four, to see if covert signaling can spread when initially rare, we
set the initial prevalence of covert signalers to be $\rho_{cov,0}=0.05$ by
initializing that fraction of agents to be covert signalers, and the rest
initialize to be overt signalers. Our outcome measure in this last case is
the number of times the prevalence of covert signaling, e.g., reaches a 
stable value above its initial value of 0.05. To ensure none of the results
from these computational experiments are not due to arbitrary settings of
what we have identified as non-explanatory parameters, we also
perform sensitivity analyses.

\subsubsection{Evolution of signaling strategies}

This basic design initializes $N$ agents with a arndomized trait vector, 
an attitude vector, and a signaling and receiving strategy.
Randomly setting agent attributes means
choosing $\pm 1$ for each element of agent trait vectors with equal 
probabilities, and choosing a signaling and receiving strategy at random
with equal probabilities for each strategy type. Attributes are initialized
according to the agent's receiving strategy. Generous agents have all 
attitudes about other agents set to 0, indicating neutrality. Churlish agents
have all attitudes set to -1, since these agents dislike others by default.
This and all experiments iterate the model $T$ times, which
means $T$ signaling rounds, with $I$ interactions between signaling rounds,
and social learning after the signaling rounds. In this basic experiment,
we analyze the global prevalence of covert signalers at time $T$ ($\rho_{cov,T}$)
and the global prevalence of generous receivers at time $T$ ($\rho_{gen,T}$),
including calculating correlation between the two.

\subsubsection{Covert signaling in minority/majority populations}
The first variation on this basic model is intended to test our 
hypothesis that covert signaling prevalence in 
minority populations ($\rho_{cov}^{minor}$) is inversely correlated with
homophily. For this, we initialize the first $M$ traits out of $K$ to mark the 
minority and majority populations. The minority population has its first
$M$ trait vector elements set to $+1$ and the majority population has its
first $M$ trait vector elements set to $-1$. For example, if $K=5$, $M=2$,
and agent $i$ is in the minority, then its trait vector would be
$\tau_i = (-1, -1, -1, \pm 1, \pm 1)$; if $i$ is in the majority, its 
trait vector would be $\tau_i = (1, 1, 1, \pm 1, \pm 1)$, where in the last
two positions the value is chosen randomly with equal probabilities.
In this experiment we analyze the prevalence of covert signalers and 
generous receivers in the minority and majority populations at time T: 
$\rho_{cov,T}^{minor}$, $\rho_{cov,T}^{major}$,
$\rho_{gen,T}^{minor}$, and $\rho_{gen,T}^{major}$.

\subsubsection{Communication strategy invasion}

To understand when a rare communication strategy can become stably established
in a population, we perform invasion analyses by setting a small fraction of
the population to have one of the four strategies, and see if that strategy
reaches a stable, increased prevalence. Whichever strategy, call it $s$,  of the invading
population, the prevalence of that strategy is set to 0.05, $\rho_{s,t=0}=\rho_{s,0}=0.05$.
When we consider strategy $s$ is invading, it is either a signaling or receiving strategy. 
Whether signaling or recieving, the non-invading part of the population has
their strategy set to the alternative for that communication process, e.g.\
if $s=cov$, then $\rho_{cov,0}=0.05$ and $\rho_{ov,0}=0.95$.

\subsubsection{Sensitivity analyses}

The main text is dedicated to demonstrating the adaptiveness of covert signaling
and churlish receiving, compared to their alternatives overt signaling and
generous receiving. There are a number of parameters that may have an outsize impact on our results
that we do not report in the main text. While we report the results of these
experiments in the Appendix, we specify which parameters we set constant for
our main results, and describe our experiments for testing model sensitivity
to these parameters

\begin{enumerate}
  \item Correlation between churlish receiving and covert signaling prevalences
  \item Introduction of $\delta \neq d$, additional penalty for both interacting
    agents disliking one another.
  \item Minority and similarity dynamics with more traits.
\end{enumerate}

@@

Here I co-vary the disliking penalties, which I set equal ($d=\delta$), and
homophily, $w$. Here, $r=0.25$ and $R=0.5$.

Specifically, I vary $d=\delta \in \{0.0, 0.05, 0.1, \ldots, 0.45\}$
As in the above experiment, I set $w \in \{0.0, 0.5, \ldots, 0.45\}$.

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/basic_disliking_signaling.pdf}
    \caption{Density of covert signalers.}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/basic_disliking_receiving.pdf}
    \caption{Density of churlish receivers.}
  \end{subfigure}

  \caption{Density of covert signalers and churlish receivers at $t=500$, 
    the final timestep.}
  \label{fig:dislikingHomophilyHeatmap}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/basic_receptivity_signaling.pdf}
    \caption{Density of covert signalers.}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/basic_receptivity_receiving.pdf}
    \caption{Density of churlish receivers.}
  \end{subfigure}

  \caption{Density of covert signalers and churlish receivers at $t=200$, 
    final timestep recorded in this preliminary experiment. Receptivity of
    overt signals is $R=0.5$.}
  \label{fig:receptivityHomophilyHeatmap}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/receptivity_allcombos_reg.pdf}
    \caption{}
    \label{fig:}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/disliking_allcombos_reg.pdf}
    \caption{}
    \label{fig:}
  \end{subfigure}
  \caption{Correlation between proportion of covert signalers and proportion of
    churlish receivers for all tested parameter combinations in both 
    receptivity (a) and disliking (b) experiments. (NEED TO CHECK HOW I'M AVERAGING
    HERE AND POSSIBLY AGGREGATE FURTHER AS A SMOOTHING FILTER. ALSO NEED TO
    )}
  \label{fig:regressions}
\end{figure}

@@

Here we analyze the evolution of covert signaling in populations where there
are minority and majority sub-populations, as specified by agents' trait
vectors, $\tau_i$. In these experiments, the first $M$ of each agent's $K$ traits
are set to 1 or -1: fraction $\rho^{minor}$ of the
population is in the minority, and have the first $M$ of their traits set to
1, and the rest of the population is the majority, with the first $M$ of their
traits set to -1.

We primarily vary the minority prevalence, $\rho^{minor}$, and set $M=1$ and
$K=3$, except for one heatmap in 
Figure~\ref{fig:covert-signaling-minority-heatmap}.

\subsubsection{Covert minority signaling over minority prevalence}

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.31\textwidth}
    \includegraphics[width=\textwidth]{Figures/minority-homophily=0p05.pdf}
    \caption{}
  \end{subfigure}
  \begin{subfigure}{0.31\textwidth}
    \includegraphics[width=\textwidth]{Figures/minority-homophily=0p25.pdf}
    \caption{}
  \end{subfigure}
  \begin{subfigure}{0.31\textwidth}
    \includegraphics[width=\textwidth]{Figures/minority-homophily=0p45.pdf}
    \caption{}
  \end{subfigure}
  \caption{}
\end{figure}

\subsubsection{Minority covert signaling over homophily and disliking}

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/covert_signalers_diff_0p05.pdf}
    % \caption{Prevalence of minority covert receivers; $\rho^{minor} = 0.10$.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/covert_signalers_diff_0p10.pdf}
    % \caption{Prevalence of minority covert receivers; $\rho^{minor} = 0.25$.}
  \end{subfigure} \\[.25in]
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/covert_signalers_diff_0p25.pdf}
    % \caption{Prevalence of minority churilsh receivers; $\rho^{minor} = 0.10$.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/covert_signalers_diff_0p10_K=10_M=4.pdf}
    \caption{(NEED TO UPDATE WITH K=9 AND M=4. Ran these, just need to pull from
    MERCED and analyze.)}
  \end{subfigure}
  \caption{Difference in covert signaling prevalence between minority and
  majority populations for different parameters settings.}
  \label{fig:covert-signaling-minority-heatmap}
\end{figure}

\subsubsection{Minority churlish receiving over homophily and disliking}

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/churlish_receivers_diff_0p05.pdf}
    % \caption{Prevalence of minority covert receivers; $\rho^{minor} = 0.10$.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/churlish_receivers_diff_0p10.pdf}
    % \caption{Prevalence of minority covert receivers; $\rho^{minor} = 0.25$.}
  \end{subfigure} \\[.25in]
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/churlish_receivers_diff_0p25.pdf}
    % \caption{Prevalence of minority churilsh receivers; $\rho^{minor} = 0.10$.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Figures/churlish_receivers_diff_0p10_K=10_M=4.pdf}
    % \caption{Prevalence of minority churilsh receivers; $\rho^{minor} = 0.25$.}
  \end{subfigure}
  \caption{Difference in churlish receiving prevalence between minority and majority
  populations for different parameter settings.}
\end{figure}

@@

In this experiment we begin with initial proportions, $\rho$ of one of the strategies
at 0.05 and consider this the ``invading'' strategy. $\rho_{cov,0}$ represents
the initial covert proportion, $\rho_{ov,0}$ the initial overt proportion,
$\rho_{ch,0}$ the initial churlish proportion, and $\rho_{gen,0}$ the 
initial generous proportion. We measure how often
a non-zero proportion of agents with the invading strategy manage to 
establish a permanent population, which we call invasion success. We calculated
invasion success for a number of experimental parameter settings, covarying
over disliking penalty ($d=\delta$) and homophily ($w$) in the first case,
and over relative covert receptivity ($r/R$) and homophily ($w$) 
in the second case (Figures~\ref{fig:cov-ov-invade} and~\ref{fig:ch-gen-invade}).

\begin{figure}[H]
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\textwidth]{Figures/covert_invades.pdf}
      \caption{Covert invades, $\rho_{cov,0}=0.05$}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\textwidth]{Figures/overt_invades.pdf}
      \caption{Overt invades, $\rho_{cov,0}=0.95$}
    \end{subfigure}
  \caption{Covert and overt sending strategy invasion success rates. 
    Lower homophily is more favorable for successful covert signaling invasion.
    This is due to lower homophily meaning agents cannot assort well into
    similar pairs prior to interaction.
    Similarly, greater homophily is more favorable for successful overt
    signaling invasion, where agents can better assort if they know each other's
    true traits. Unexpectedly, as covert receptivity increases, overt signaling
    is is more successful at invading and covert signaling is less successful
    at invading when the initial churlishness is high. When initial
    churlishness is small to moderate, the pattern is reversed.
  }
  \label{fig:cov-ov-invade}
\end{figure}

\begin{figure}[H]
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\textwidth]{Figures/churlish_invades.pdf}
      \caption{Churlish invades, $\rho_{ch,0}=0.05$}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\textwidth]{Figures/generous_invades.pdf}
      \caption{Generous invades, $\rho_{ch,0}=0.95$}
    \end{subfigure}
  \caption{Churlish and generous receiving strategy invasion success rates.
    For low to moderate rates of initial covert signaling there is not much
    of a pattern to invasion success. High values of homophily do seem to lead
    to more successful churlish invasion for moderate values of initial
    covert signalers. For large initial proportions of covert
    signalers, lower homophily favors generous invasion, which is more pronounced
    for larger disliking penalties.
  }
  \label{fig:ch-gen-invade}
\end{figure}

\subsubsection{Non-signaling as an adaptive strategy}

One of the most striking patterns in the invasion results is the evolution
of a third de facto strategy of non-signaling. This strategy emerges when
we consider both the invasion dynamics of either the covert or overt 
signaling strategy. When simulating the invasion of covert signaling
($\rho_{cov,0}=0.05$), the non-signaling strategy emerges when the covert
signaling efficiency, $r$, is 0, but covert signaling still successfully
invades as the chosen agent strategy. When $r=0$, no agents receive
covert signals, making this more accurately a \emph{non-signaling} strategy.  
For covert signaling invasion, this only happens when the initial churlishness
is low ($\rho_{ch,0}=0.05$) and homophily is also low ($w \lessapprox 0.2$). 
Non-signaling wins out evolutionarily 
in some cases when considering of invading overt signalers when overt signalers
fail to invade but $r=0$, again meaning agents in this society prefer not
to signal at all. This occurs some of the time for all tested homophily
values and low initial churlishness ($\rho_{ch,0}=0.05$) (Figure~\ref{fig:cov-ov-invade}b, bottom left plot).  
Overt signaling continues to fail to evolve at least some of the time
for $r=0$ for moderate initial churlishness ($\rho_{ch,0}=0.5$) up up to perfect 
assortment homophily, $w=0.5$.
